Hybrid Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-10-01 manuscript/enar/enar_cls.tex
basis approximation

>>>>

\section{Hybrid Hilbert Space}
Our framework leverages that the nonlinear iterative PLS (NIPALS; \citealp{woldPathModelsLatent1975}) algorithm, a standard PLS fitting method, operates solely via inner-product space operations. We define a Hilbert space for hybrid predictors with a complete inner product to extend the algorithm.
%
\begin{theorem}[Hybrid space]\label{def:hilbert_space}
	An element $h \in \mathbb{H} := (\mathbb{L}^2[0,1])^K \times \mathbb{R}^p$ is an ordered tuple $h = (f_1, \dots, f_K, \mathbf{u})$, where $f_k \in \mathbb{L}^2[0,1]$ for $k=1, \dots, K$ and $\mathbf{u} \in \mathbb{R}^p$.  
We define an inner product on $\mathbb{H}$ for any two elements $h_1 = (f_1, \dots, f_K, \mathbf{u})$ and $h_2 = (g_1, \dots, g_K, \mathbf{v})$ as:
\begin{equation}\label{eq: hybrid inner product}
	\langle h_1, h_2 \rangle_{\mathbb{H}} := \sum_{k=1}^K \int_0^1 f_k(t) g_k(t) \, dt +   \mathbf{u}^\top \mathbf{v}.
\end{equation}
	The inner product induces a norm $\Vert \cdot \Vert_{\mathbb{H}}$ on the space, defined as $\Vert h \Vert_{\mathbb{H}} := \langle h, h \rangle_{\mathbb{H}}^{1/2}$, and a corresponding metric $d(h_1, h_2) = \Vert h_1 - h_2 \Vert_{\mathbb{H}}$.
	Then the hybrid inner product space $\mathbb{H}$ is a separable Hilbert space.
\end{theorem}

\begin{definition}[Hybrid Predictor]\label{def:hybrid_predictor}
	For the Hilbert space  $\mathbb{H}$ 
	defined in Definition \ref{def:hilbert_space},
	The Borel $\sigma$-field on $\mathbb{H}$, denoted $\mathfrak{B}(\mathbb{H})$, is the smallest $\sigma$-field containing the class $\mathfrak{M}$ of all sets of the form
	$
	\{q \in \mathbb{H} \mid \langle q, h \rangle \in O\},
	$
	for any $h \in \mathbb{H}$ and any open subset $O \subseteq \mathbb{R}$ (details can be found in Theorem 7.1.1 of \citealt{hsingTheoreticalFoundationsFunctional2015}).
	A hybrid predictor $ W_i = (X_{i1}(t), \ldots, X_{iK}(t), \mathbf{Z}_i)$ is a measurable mapping from a probability space $(\Omega, \mathfrak{F}, P)$ into   $
	\bigl(
	\mathbb{H}, \mathfrak{B}(\mathbb{H})
	\bigr)
	$. 
\end{definition}
Then the joint regression model \eqref{eq:hybrid_regression_model}  can be concisely written as
\begin{equation}		\label{eq: Hybrid functional model}
	Y_i = \langle \beta, W_i \rangle_{\mathbb{H}} + \epsilon_i,~\text{where}~\beta := \bigl( \beta_1(t), \ldots, \beta_K(t), \boldsymbol{\beta} \bigr)\in \mathbb{H}.
\end{equation}

\section{Proposed PLS Algorithm}\label{section:main:our_algorithm}
Our approach efficiently computes PLS components and scores for multiple dense or irregular functional predictors alongside scalar predictors.   It incorporates regularization to exploit structural relationships within and between the functions, to prevent overfitting and improve interpretability.
Each iteration involves two subroutines: regularized component estimation (Section \ref{section:sub:compute_PLS_component})  and residualization 
(Section \ref{section:sub:residualization})
After iteration terminates 
the hybrid regression coefficient is estimated (Section~\ref{section:sub:regression_coeff}).

\subsection{Preliminary step 1: finite-basis approximation}\label{sec: finite basis approximation}
Let $\{b_m(t)\}$ be a twice-differentiable basis of $\mathbb{L}^2([0,1])$ whose second derivatives are also linearly independent, for example, cubic B-splines, Fourier, or orthonormal cubic polynomials. 
Fourier coefficients of
functional predictors
$X_{ij}(t)$,
functional regression coefficients
$\beta_j(t)$,
PLS directions,
$\xi_j(t)$,
and residualization coefficients
$\delta_j(t)$
(with iteration indices suppressed) with respect to 
$\{b_m(t)\}$ are denoted as
$\{\theta_{ijm}\},
\{ \eta_{jm} \},
\{ \gamma_{jm} \},
$
and
$\{ \pi_{jm} \}$.
For practicality, we truncate the expansion at a moderate $M$ (e.g., 15-20) to capture most functional variation, while smoothness is handled via penalization (Section~\ref{sec: Regularization of PLS Components}).
The truncated expansion for the functional data is denoted
as
are denoted as $\widetilde{X}_{ij}(t)
:=
\sum_{m=1}^M \theta_{ijm}  b_m(t)$, but for the other funtions, we denote
\begin{equation}\label{def:truncated_zeta_and_delta}
	\beta_j(t)
	:=
	\sum_{m=1}^M \eta_{jm} b_m(t),
	%
	\xi_j(t) := \sum_{m=1}^M \gamma_{jm}  b_m(t),
	%
	\delta_j(t)
	=
	\sum_{m=1}^M
	\pi_{jm}
	b_m(t),
\end{equation}
since these functions are estimated, using a hat over the tilde notation would be cumbersome.     Let Let $\boldsymbol{\theta}_{ij}$, $\boldsymbol{\eta}_j$, $\boldsymbol{\gamma}_j$, and $\boldsymbol{\pi}_j$ denote these coefficients as $M$-dimensional vectors.
The truncations imply that all computations in this paper are carried out entirely within the subspace
\begin{equation}\label{def:truncated_hilbert_space}
	\widetilde{ \mathbb{H} } := \operatorname{span}\bigl(b_1(t), \ldots, b_M(t)\bigr)^K \times \mathbb{R}^p \subset \mathbb{H}.
\end{equation}
The $i$th hybrid predictor, projected on $\widetilde{ \mathbb{H} }$, is represented by the tuple
$ 
\widetilde{W}_i := 
(
\widetilde{X}_{i1}, \ldots, \widetilde{X}_{iK}, \boldsymbol{Z}_i
).
$ 
For notational convenience, we stack the predictor coefficient vectors as
\begin{equation}\label{def:theta}
	\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M},
	\quad 
	\Theta := (\Theta_1, \ldots, \Theta_K, \mathbf{Z}) \in \mathbb{R}^{n \times (MK + p)}.
\end{equation}
Let us denote the response vector as $\mathbf{y} := (y_1, \ldots, y_n)^\top$.
Let $B, B^{\prime\prime} \in \mathbb{R}^{M \times M}$ denote the Gram matrices of the basis functions and their second derivatives, with entries
\begin{equation}\label{def:B_matrix}
	B_{m, m'} := \int_0^1 b_m(t)\, b_{m'}(t)\, dt, \quad
	B^{\prime\prime}_{m, m'} := \int_0^1 b_m''(t)\, b_{m'}''(t)\, dt,
\end{equation}
for $m, m' = 1, \ldots, M$. We then define the block-diagonal matrices
\begin{equation}\label{def:gram_block}
	\mathbb{B} := \operatorname{blkdiag}(B, \ldots, B, I_p), \quad
	\mathbb{B}^{\prime\prime} := \operatorname{blkdiag}(B^{\prime\prime}, \ldots, B^{\prime\prime}, I_p),
\end{equation}

Then the full data for the hybrid PLS problem at the $l$-th iteration can be represented by the tuple
\begin{equation}\label{def:problem_instance}
	(\mathbb{B}, \mathbb{B}^{\prime\prime}, \Theta, \mathbf{y}) \in 
	\mathbb{R}^{(MK+p) \times (MK+p)} \times 
	\mathbb{R}^{(MK+p) \times (MK+p)} \times 
	\mathbb{R}^{n \times (MK+p)} \times 
	\mathbb{R}^n,
\end{equation}
with the index $l$ omitted for brevity.

\begin{remark}
	Different bases could be used for each functional predictor to handle  multiple dense or irregular functional predictors. We adopt a common basis for simplicity. The definitions of $\mathbb{B}$ and $\mathbb{B}^{\prime\prime}$ remain general enough to accommodate distinct bases if needed.
\end{remark}
